---
title: "Bootcamp Final Exercise"
author: "Ruyi Lu"
date: "2024-09-23"
output: html_document
---

### Questions to look at
#### What's the difference in test performance between low, medium and high poverty areas? Has this relationship changed over time? Is this relationship at all moderated by access to free / reduced price lunch?

```{r setup, include=FALSE}
library(tidyverse)
```

### load data
```{r}
acs = read_csv('data/nys_acs.csv')
school = read_csv('data/nys_schools.csv')
#acs: 496*5
#school: 35663*12
```

### data cleaning
```{r}
sum(is.na(acs))
# no missing values in acs

sum(is.na(school))
# 1706 missing values in school
# but NA labeled as -99, convert -99 to na first

str(school)
# colnames(school)

# divide school by variable type, numeric and character
# drop `district_name`
num = c("year","total_enroll","per_free_lunch","per_reduced_lunch","per_lep","mean_ela_score","mean_math_score")
chr = c("school_cd","school_name","county_name", "region","year")
school_num = school[,num]
school_chr = school[,chr]

# convert -99 in numerical variables to NA
school_num=replace(school_num, school_num==-99, NA)
colSums(is.na(school_chr))
colSums(is.na(school_num))

#replace na with mean
school_num$total_enroll=ifelse(is.na(school_num$total_enroll), mean(school_num$total_enroll, na.rm = TRUE), school_num$total_enroll)
school_num$per_free_lunch=ifelse(is.na(school_num$per_free_lunch), mean(school_num$per_free_lunch, na.rm = TRUE), school_num$per_free_lunch)
school_num$per_reduced_lunch=ifelse(is.na(school_num$per_reduced_lunch), mean(school_num$per_reduced_lunch, na.rm = TRUE), school_num$per_reduced_lunch)
school_num$per_lep=ifelse(is.na(school_num$per_lep), mean(school_num$per_lep, na.rm = TRUE), school_num$per_lep)

colSums(is.na(school_num))

# scale math and ela score
school_num$mean_ela_score=scale(school_num$mean_ela_score)
school_num$mean_math_score=scale(school_num$mean_math_score)

# combined school dataset together
# drop repetitive column `year.1`
temp = cbind(school_chr, school_num)
str(temp)
temp[,6] = NULL

# dont wanna use rf or mice to replace NA so will just drop them here
school_cleaned = na.omit(temp)

colSums(is.na(school_cleaned))
```

### poverty level seperation
```{r}
# categorical variable grouping counties into "high", "medium", and "low" poverty groups
# split based on 33% and 67% percentile
str(acs)
fivenum(acs$median_household_income)
quantile(acs$median_household_income, c(.01, .33, .67, .99)) 
acs$poverty_level<-NA
acs$poverty_level <- ifelse(acs$median_household_income < 47627.80, 'high',
                            ifelse(acs$median_household_income <= 53488.60, 'medium', 'low'))
```

### merge data
```{r}
temp <- school_cleaned %>% left_join(acs, by=c("county_name","year"))
data = na.omit(temp)
colSums(is.na(data))
```

### analysis
### summary table for each county: total enrollment, percent of students qualifying for free or reduced price lunch, and percent of population in poverty.
```{r}
# total enrollment by county
data %>%
  group_by(county_name) %>%
  summarize(total_enrollment = sum(total_enroll))

# average per_free_lunch by county
data %>%
  group_by(county_name) %>%
  summarize(avg_per_free_lunch = mean(per_free_lunch))

# average per_reduced_lunch by county
data %>%
  group_by(county_name) %>%
  summarize(avg_per_reduced_lunch = mean(per_reduced_lunch))

# general poverty level
table(data$poverty_level)

# poverty level of each county
data %>%
  group_by(county_name, poverty_level) %>%
  summarize(count = n(), .groups = 'drop') %>% head()

# general poverty level over year
data %>%
  group_by(year, poverty_level) %>%
  summarize(count = n(), .groups = 'drop') %>% head()

```

### my own analysis
```{r}
data %>% group_by(county_name) %>% arrange(county_per_poverty) %>% head(10)

# average math score by poverty level for each county?
data %>% group_by(county_name, poverty_level) %>%
  summarise(avg_math_score = mean(mean_math_score)) %>% head(10)

# higher poverty rate counties/regions -> lower score performance?
# low-income neighborhoods < high-income neighborhoods in terms of score and percentage of bach?
# relationship btw poverty and test performance over years?

# which is better measure of household income, math score or ela score?
ela = lm(median_household_income~mean_ela_score, data)
math = lm(median_household_income~mean_math_score, data)
summary(ela)
summary(math)
# neither seems like a good measurement intuitively
# will use math score anyway

```

### viz
```{r}
# over the past few years, the difference in percentage of bach?
data %>% 
  group_by(year, poverty_level) %>%  # Include year in grouping
  summarise(avg_county_bach_perc = mean(county_per_bach)) %>% 
  ggplot(aes(x = year, y = avg_county_bach_perc, group = poverty_level, color = poverty_level)) + 
  geom_smooth() + 
  labs(title = "Percenatge of Bachelor of Different Poverty Level over Year",
       x = "Year", y = "Percentage of Bachelor") + 
  theme(legend.position = "bottom")


# over the past few years, the difference in percentage of bach between low, medium and high poverty areas?
data %>% 
  group_by(year, poverty_level) %>%
  summarise(avg_county_bach_perc = mean(county_per_bach)) %>%
  ggplot(aes(x=poverty_level, y=avg_county_bach_perc, fill = poverty_level)) +
  geom_bar(stat = 'identity') + 
  facet_wrap(~year) + 
  labs(title="Percenatge of Bachelor of Different Poverty Level over Year",
       x="Poverty Level", y="Percenatge of Bachelor") + 
  theme(legend.position = "bottom")


```

