---
title: "Intro to R: Loops, Conditionals, Functions"
author: "Richard Paquin Morel"
date: "`r format(Sys.Date()+1, '%B %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
params:
  notes: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4.35, fig.width = 4.75)
```

```{r, include=FALSE}
notes<-params$notes
```

## Remember our data from Monday...

```{r revisiting the data}
gapminder <- read.csv(here::here("data", "gapminder5.csv"))
head(gapminder)
```

## Sidebar: the `here` package

The `here()` function gets your current working directory and appends any strings enclosed in the function.

```{r here}
# install.packages("here")
here::here()

here::here("data", "gapminder.csv")
```

## Remember our data from Monday...

```{r revisiting the data 2}
gapminder <- read.csv(here::here("data", "gapminder5.csv"))
head(gapminder)
```


## `gapminder` dataset

```{r str gapminder}
str(gapminder)
```

## `gapminder` dataset

- The dataset has data on 142 countries on 5 continents for five year intervals between 1952 and 2007
- Let's fix a few things about the data...the character variables are currently factors. Doesn't make sense!
- Find the factor variables and change them to character vectors

## `gapminder` dataset

```{r recoding gapminder}
gapminder$country <- as.character(gapminder$country)
gapminder$continent <- as.character(gapminder$continent)
str(gapminder)
```

## `gapminder` dataset
- What if we want to know the mean life expectancy by country? 
  -Or by year?
- One approach:
```{r one approach}
mean(gapminder$lifeExp[gapminder$country == "Afghanistan"])
mean(gapminder$lifeExp[gapminder$country == "Albania"])
#...
```

## `gapminder` dataset
- Or we could do save ourselves a lot of typing and time

- Loops!

# Don't repeat yourself--loop it! 

## Loops in R -- `for`, `while`, and the `apply` family

- Loops allow you to repeat a function based on conditions
- There are several flavors
  - `for`
  - `while`
  - the `apply` family (often preferred to loops because the code is cleaner)
- Sneak peak: `dplyr` and `data.table` offer better approaches to loops

## A word of caution about loops

- It is often a good idea at avoid loops when possible in R
  - They are slow!
- Try to find other solutions before looping

## The anatomy of a `for` loop

- `for` loops repeat a function for all values in a vector  -- _don't cut and paste!_
- The basic form is: `for (i in vector) { function(i) }`
  - `i` is the iterator variable (could be any letter!)
  - The loop repeats for each value in the vector, which replaces `i` for each interation
  
## The anatomy of a `for` loop

- Let's recover the GDP for each country

```{r loop example}
# create a vector of values that you want to repeat the function for
obs <- 1:nrow(gapminder)

# initialize the for loop with `for (i in vector)` 
for (i in obs) { # the function to repeat is enclosed in braces {}
  gapminder[i, "gdp"] <- gapminder[i, "pop"] * gapminder[i, "gdpPercap"]
}
```

## You try

Create a new variable that finds that natural log (`log`) of the GDP per capita and of population
 - call them `log_gdpPercap` and `log_pop`

## Log gdp and log population

```{r loop exercise}
# initialize the for loop with `for (i in vector)` 
for (i in obs) { # the function to repeat is enclosed in braces {}
  gapminder[i, "log_gdpPercap"] <- log(gapminder[i, "gdpPercap"])
  gapminder[i, "log_pop"] <- log(gapminder[i, "pop"])
}
```


## Log gdp and log population

```{r log exercise}
head(gapminder)
```

## Avoid loops when possible

- Loops are useful, but slow
- Avoid when possible, especially when there is a vectorized function you can use

```{r vectorized}
gapminder$vec_log_gdpPercap <- log(gapminder$gdpPercap)
all(gapminder$vec_log_gdpPercap == gapminder$log_gdpPercap)
```


## Let's try something a bit more substantive

- Has life expectancy increased over time?
- Find the mean life expectancy by year

```{r mean by year}
years <- unique(gapminder$year)

for (i in years) {
  mean_le <- mean(gapminder$lifeExp[gapminder$year == i], 
                  na.rm = T)
  print(paste0(i, ": ", mean_le))
}
```

## Mean life expectancy by continent

- Try the same, this time for continents!
  - Which continent has the highest mean life expectancy?

## Mean life expectancy by continent
```{r mean by continent}
conts <- unique(gapminder$continent)

for (i in conts) {
  mean_le <- mean(gapminder$lifeExp[gapminder$continent == i], 
                  na.rm = T)
  print(paste0(i, ": ", mean_le))
}
```

## double up!

- It is possible to make nested `for` loops by defining different iterators
- What is the mean life expectancy for each continent for each year?

## the nested `for` loop 

```{r nested for loop}
for (i in conts) {
  print(paste0("Continent: ", i))
  for (j in years) {
    mean_le <- mean(gapminder$lifeExp[gapminder$continent == i & 
                                          gapminder$year == j], 
                      na.rm = T)
    print(paste0(j, ": ", mean_le))
  }
  }
```

## nested `for` loop exercise!

- Has the gap in life expectancy between countries on different continents narrowed over time?

## nested `for` loop exercise!

- What is the standard deviation (`sd`) for life expectancy for each continent for each year?

## nested `for` loop exercise!

```{r nested for loop exercise!}
for (i in conts) {
  print(paste0("Continent: ", i))
  for (j in years) {
    sd_le <- sd(gapminder$lifeExp[gapminder$continent == i & 
                                          gapminder$year == j], 
                    na.rm = T)
    print(paste0(j, ": ", sd_le))
  }
}
```

# `for` loops can be slow...very slow

## `for` loops can be slow...very slow

- Sometimes, people recommend using the `apply` family of functions as a faster alternative
  - There are _some_ efficiency gains with smaller datasets
  - But, as Patrick Burns, the author of "R Inferno", states, the `apply` family is "loop-hiding"
- `apply` and its relatives help you write cleaner code, but do not expect much of a speed boost
  - do a search for "for loops versus apply in r" and get a taste for the debate

## Three flavors: `apply`, `lapply`, `sapply`

- Let's look at these three (there are more!): `apply`, `lapply`, `sapply`

## `apply`

`apply(matrix, 1 = row or 2 = column, function)`
- Let's say we want to find the mean for each stat in `gapminder`

```{r apply}
vars <- gapminder[, c("lifeExp", "pop", "gdpPercap")]
apply(vars, 2, mean)
```

## `apply` versus `for`

```{r apply vs for}
apply(vars, 2, mean)

for (i in vars) {
  print(mean(i))
}
```

## `lapply` and `sapply`

- Both `lapply` and `sapply` iterate over a values in a vector or list, rather than rows or columns
  - Generally, much more common to use in data analysis
- `lapply` returns a list
- `sapply` returns a simplified list (i.e., a vector)
  - Word of caution -- there is some inconsistency in how `sapply` returns results, so always check

## `lapply` and `sapply`

- `lapply(vector, function)`

```{r lapply, warning = F, message = F}
lapply(gapminder, mean)
sapply(gapminder, mean)
```

## Anonymous functions in `apply`

- You can do more complex functions within an `apply` call
- Add `function(x) [function]` to the call--`x` becomes the iterator

```{r anonymous function}
sapply(years, function(x) mean(gapminder$lifeExp[gapminder$year == x]))
```

# the `while` loop

- The `for` loop repeats a function for all values in vector
- But what if we want to repeat a function until a condition is no longer met?
  - Say hi to the `while` loop!
  - For example, what is the standard deviation of life expectancy for years before 1987?

## `while` loop syntax

- Similar syntax to the `for` loop -> `while (condition) { function }`
- Often, you define an interator that you will increase for each loop

```{r while loop}
i <-  1952 # define the interator

while (i < 1987) {
  sd_lf <- sd(gapminder$lifeExp[gapminder$year == i])
  print(paste0(i, ": ", sd_le)
        )
  i <- i + 5 # increase the iterator by the interval between years
}
```

## try a `while` loop

- What is the standard deviation for life expectancy for each year between 1987 and 2002 (inclusive)?

## results

```{r while loop exercise}
i <-  1987 # define the interator

while (i <= 2002) {
  sd_lf <- sd(gapminder$lifeExp[gapminder$year == i])
  print(paste0(i, ": ", sd_le)
        )
  i <- i + 5 # increase the iterator by the interval between years
}
```

## the infinite loop - a `while` loop cautionary tale

- Beware! A `while` loop will continually run if the logical condition is always satisfied!
- Give it a try: run the previous `while` loop _without increasing the iterator_

```{r the infinite loop, eval = F}
i <-  1987 # define the interator

while (i <= 2002) {
  sd_lf <- sd(gapminder$lifeExp[gapminder$year == i])
  print(paste0(i, ": ", sd_le)
        )
  # oops! forgot to increase i by 5! ESC! ESC! ESC!
}
```

# The `if/else` conditional!

- `if/else` - executes a function when a condition is met

## Start with `if`

- Similar to `for` and `while`, initialize with `if` and then detail condition in parentheses

```{r if statement}
random_year <- sample(years, 1)

random_year

if (random_year > 1977) {
  print(random_year)
}
```

## `if` statements

- We probably got different answers because of random sampling from `years`
- We can fix that with `set.seed()`

```{r if statement with set seed}
set.seed(10)
random_year <- sample(years, 1)

if (random_year > 1977) {
  print(random_year)
}
```

## Adding an `else` clause

- What happens if `random_year` <= 1977? NOTHING!
- We can add an `else` statement, telling R what to do when the `if` condition isn't met

```{r adding the else statement}
set.seed(1)
random_year <- sample(years, 1)
random_year

if (random_year > 1977) {
  print(random_year)
  } else {
  print("sorry, random year is less than 1977")  
  }
```

## putting `if` and `else` together

```{r if else statement}
random_year <- sample(years, 1)

if (random_year > 1977) {
  print(paste0(random_year, ": ", 
               mean(gapminder$lifeExp[gapminder$year == random_year]))
        )
  } else { 
    print("sorry, random year is less than 1977")
    }
```

## Putting `for` and `if/else` together

- We can add an `if`...`else` clause to a `for` loop

```{r, eval = F}
# initialize the `for` loop
for () {
  if () { #initialize if statement
    # function if condition is met
  } else {
    # function if condition is  not met
  }
}
```


## Putting `for` and `if/else` together

Which continents have a mean life expectancy greater than 70 years?

```{r for loop with if else}
threshold <- 70

for (i in unique(gapminder$continent)) {
   tmp <- mean(gapminder$lifeExp[gapminder$continent==i])
   
   if (tmp < threshold) {
       print(paste("Mean Life Expectancy in", i, "is less than", threshold))
   } else {
       print(paste("Mean Life Expectancy in", i, "is greater than", threshold))
   }
}
```

## Putting `for` and `if/else` together

Write a `for` loop that reports the mean population for years greater than or equal to 1987. Make sure the loop prints a message if the condition is not met!

## Putting `for` and `if/else` together

```{r for if else exercise}
for (i in years) {
  if (i >= 1987) {
    mean_pop <- mean(gapminder$lifeExp[gapminder$year == i])
    print(paste0(i, ": ", mean_pop))
  } else {
    print("Sorry, year is less than 1987")
  }
}
```


# Writing functions

## Writing functions

- Hadley Wickham: If you have to copy-and-paste three times, it is time to write a function
- John Chambers: Everything that happens in R is the result of a function call -- even `[`

```{r everything is a function call}
gapminder$lifeExp[gapminder$country == "Germany"]
`[`(gapminder$lifeExp, gapminder$country == "Germany")
```

## The anatomy of a function

- Write a function should look familiar -- initialize with `function`
- Every function has arguments

```{r a function, eval = F}
my_function <- # give the function a name
  function(x, y) { # arguments for the function go inside the parentheses
    # the expressions do in the braces
  }
```

## Writing a simple function

- Let's write a simple function that prints the value of a selected variable in the gapminder dataset

```{r a simple function}
get_values <-
  function(df, variable = "continent") {
    vals <- unique(df[[variable]])
    print(paste0(variable, ": ", vals))
  }
```

## Writing a more substantial function

- Let's write a function that prints the mean and standard deviation for life expentancy for a given country in the `gapminder` dataset

```{r writing a function}
report_mean_sd <- 
  function(df, variable, country) {
    var <- df[[variable]][df$country == country]
    m_le <- mean(var)
    sd_le <- sd(var)
    cat("Country:", country, 
        "\nMean Life Expectancy:", m_le,
        "\nSD Life Expectancy:", sd_le)
  }

report_mean_sd(gapminder, "lifeExp", "Bulgaria")
```


## Create your own function

- Write a function that reports the mean, median, minimum, and maximum for life expectancy for a continent in `gapminder`
- Hint: `min`, `max`

## Create your own function

```{r your function}
report_stats <-
  function(df, variable, continent) {
    var <- df[[variable]][df$continent == continent]
    min_le <- min(var)
    max_le <- max(var)
    cat("Continent:", continent,
        "\nMinimum Life expectancy:", min_le,
        "\nMaximum Life expectancy:", max_le)
  }

report_stats(gapminder, "lifeExp", "Asia")
```


## Functions and loops, together at last

- Combining functions and loops saves time

## A log-log model relating life expectancy to GDP

```{r another function}
viz_lm <-
  function(df, dv, iv, year) {
    dat <- df[df[["year"]] == year, ]
    y <- log(dat[[dv]])
    x <- log(dat[[iv]])
    fit <- lm(y ~ x)
    plot(y ~ x, main = year,
         xlab = iv, ylab = dv)
    lines(x, predict(fit), col = 'blue')
  }
```

## Running the function

```{r running the function}
viz_lm(gapminder, "lifeExp", "gdpPercap", 1977)
```


## Loop it!

```{r}
for (i in years) {
  viz_lm(gapminder, "lifeExp", "gdpPercap", i)
}
```

# Reporting analysis with Rmarkdown and GitHub

## Rmarkdown & GitHub

- Rmarkdown creates dynamic reports in HTML, PDF, and Word
- Combine text (using the markdown language) and R code
- Rmarkdown runs R code, compiles, and produces a report in chosen format
- This presentation was created using Rmarkdown!

## R4 Exercise

1) Open the `rmd_exercise_template.Rmd`
2) Save as: `R4RExercise_LastnameFirstname.Rmd`.
2) Read in the gapminder data set
3) As you answer questions, be sure to annotate your work with as much detail as possible!
